Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2022-06-30 sports_analytics.Rmd
team strength: BT example

```{r setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  fig.align = 'center'
)
library(tidyverse)
```

@@

@Lock2014 used random forests as a means for estimating win probability before each play in a football game. Covariates included in this win probability model are the current down, score differential, time remaining, adjusted score, point spread, number of timeouts remaining for each team, total points scored, current yardline, and yards to go for a first down.

In addition, @Yurko2019 estimated win probability in the NFL using a generalized additive model (GAM), as part of the `nflscrapr` package and nflWAR framework. This model takes into account the estimated expected points obtained from the model described in \ref{sec:ev}, along with other predictors for time, current half, and timeouts. A major highlight of this framework is it is fully reproducible and uses publicly available data. One of @Yurko2019's goals was also to encourage researchers to "use, explore, and improve upon our work", which ultimately inspired \pkg{nflfastR}, known as the successor to `nflscrapr`.

These two frameworks were also implemented in @Yam2019 with minimal modifications. Specifically, the authors combined both @Lock2014 and @Yurko2019's approaches to estimate the win probability for each play, with an overall goal of assessing fourth down decision making in gridiron football.

(add a wp example from nflfastR or nflscapr)

@@

@Stern1994 used Brownian motion to investigate the scoring process of basketball. Let $p(l, t)$ represents the win probability for the home team given an $l$-point lead after $t$ seconds of game time. The model introduced by @Stern1994 is a probit regression model, which provides an estimate for $p(l, t)$. In particular,

$$
p(l, t) = \Phi\left(\frac{l + (1-t)\mu}{\sqrt{(1-t)\sigma^2}}\right)
,
$$

Here, a Brownian motion process with drift $\mu$ points advantage for the home team and variance $\sigma^2$ is used to model the score difference between the home and away teams.

@@

Examples of win probability can also be found in other sports, with a diverse range of statistical techniques being used to estimate the probability of winning for a player or team.

@Brenzel2019 used a three-dimensional Markov model to estimate win probability in curling.

In esport, @Maymin2020 proposed a well-calibrated in-game win probability model for each specific moment during a game of League of Legends.

More recently, @Guan2022 developed an in-game win probability model for the National Rugby League using a functional data analysis approach.

@@

A popular method for estimating team strength in sports is through pairwise evaluations. Perhaps the most famous probability model for predicting the outcome of a paired comparison is the Bradley-Terry model (BTM) [@Bradley1952]. For a pair of players (or teams) $i$ and $j$, let $\Pi_{ij}$ denote the probability that $i$ is preferred to $j$. Then the BTM is a logistic regression model with parameters $\beta_i, \beta_j$ such that

@@

Here, $\exp{\beta_i}$ is often viewed as a representation of player $i$'s ability. The BTM can be implemented in `R` [@Rcitation2022] via the \pkg{BradleyTerry2} package [@R-BradleyTerry2].

For example, ...

```{r btm, echo=TRUE}
library(BradleyTerry2)
# show BTM for Pirates vs. Yankees in 1960
# or Red Sox vs. Yankees in 2003/2004
library(retrosheet)
games <- get_retrosheet("game", 2004)

@@

btm2 <- btm |>
  update(formula = ~ team + at_home, data = bt_games2)
btm2
```

```{r}
btm |>
  broom::tidy() |>
  arrange(desc(estimate))
btm2 |>
  broom::tidy() |>
  arrange(desc(estimate))
```

Another widely known tool for measuring team strength is the Elo rating system [@Elo1978]. The Elo rating was originally developed for chess, but also has been used to estimate team strength in other sports.

See @koning2017rating for more information on applications of the Elo rating in soccer.

@Glickman1998 proposed a Bayesian state-space model for paired comparisons for predicting NFL games, allowing team strength parameters to vary over time. In particular, this paper modeled point differential in the NFL by introducing week-to-week and season-to-season as the two primary sources variation in team strengths. See also @glickman2017estimating for more discussion on estimating team strengths in American football.

@Lopez2018 extended @Glickman1998's state-space model to understand randomness in the four major American sports leagues. Betting moneylines were used in place of point differentials in order to estimate team strengths, and this framework also accounts for home advantage.

>>>>

```{r setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = 'center'
)
library(tidyverse)
```

@@

@Lock2014 used random forests as a means for estimating win probability before each play in a football game. Covariates included in this win probability model are the current down, score differential, time remaining, adjusted score, point spread, number of timeouts remaining for each team, total points scored, current yardline, and yards to go for a first down. In addition, @Yurko2019 estimated win probability in the NFL using a generalized additive model (GAM), as part of the `nflscrapr` package and nflWAR framework. This model takes into account the estimated expected points obtained from the model described in \ref{sec:ev}, along with other predictors for time, current half, and timeouts. The two win probability frameworks proposed by @Lock2014 and @Yurko2019 were also implemented in @Yam2019 with minimal modifications. Specifically, the authors combined both approaches to estimate the win probability for each play, with an overall goal of assessing fourth down decision making in gridiron football.

A vital highlight of @Yurko2019's win probability model is that it is fully reproducible and uses publicly available data. One of @Yurko2019's goals was also to encourage researchers to "use, explore, and improve upon our work", which ultimately inspired \pkg{nflfastR}, known as the successor to `nflscrapr`. Figure \ref{fig:nflfastr_wp} shows a win probability graph for the 2017-18 NFL Playoffs Divisional Round matchup between the New Orleans Saints and the Minnesota Vikings. The chance of winning for each team is obtained using the \pkg{nflfastR} R package, which implemented a gradient boosting model via the `xgboost` library for estimating win probabilities. Minnesota was leading throughout the first three quarters of the game, having win probabilities of 0.869, 0.941, 0.742 at the end of the first, second, and third quarters, respectively. The win probabilities get close to parity late in the fourth quarter, when the Saints took the lead with 3:01 seconds left in the game. The last play of this game, famously known as the Minneapolis Miracle, resulted in a drastic swing in win probabilities for both teams. With 10 seconds remaining of game time, the Vikings begin the final possession with a 25.3% chance of winning. Their probability increased to a perfect 1 when Stefon Diggs scored a game-winning 61-yard receiving touchdown as the game clock expired.

```{r}
# win prob for Minneapolis Miracle
library(teamcolors)

vikings <- teamcolors |> 
  filter(league == "nfl", name == "Minnesota Vikings") |> 
  pull(primary)

saints <- teamcolors |> 
  filter(league == "nfl", name == "New Orleans Saints") |> 
  pull(primary)
```

```{r nflfastr_wp, fig.cap="Win probability graph for Saints vs Vikings in the 2017-18 NFL Playoffs."}
library(nflfastR)
load_pbp(2017) |> 
  filter(season_type == "POST",
         home_team == "MIN",
         away_team == "NO",
         !is.na(home_wp),
         !is.na(away_wp)) |>
  select(game_seconds_remaining, home_wp, away_wp) |> 
  pivot_longer(!game_seconds_remaining,
               names_to = "team",
               values_to = "wp") |>
  ggplot(aes(game_seconds_remaining, wp, color = team)) +
  geom_line(size = 1.5) +
  scale_x_reverse(breaks = seq(0, 3600, 900)) + 
  scale_color_manual(values = c(saints, vikings), 
                     labels = c("New Orleans Saints", "Minnesota Vikings")) +
  # geom_hline(yintercept = 0.5, color = "gray", linetype = "dashed") +
  # geom_vline(xintercept = seq(0, 2700, 900), linetype = "dashed", color = "gray") + 
  labs(x = "Time Remaining (seconds)",
       y = "Win Probability",
       color = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

@@

@Stern1994 used Brownian motion to investigate the scoring process of basketball. Let $p(l, t)$ represents the win probability for the home team given an $l$-point lead after $t$ seconds of game time. The model introduced by @Stern1994 is a probit regression model, which provides an estimate for $p(l, t)$. In particular, 
$$
p(l, t) = \Phi\left(\frac{l + (1-t)\mu}{\sqrt{(1-t)\sigma^2}}\right)
,
$$
Here, a Brownian motion process with drift $\mu$ points advantage for the home team and variance $\sigma^2$ is used to model the score difference between the home and away teams.

@@

The idea of win probability is also applied to other sports, with a diverse range of statistical techniques being used to estimate the probability of winning for a player or team. @Brenzel2019 used three-dimensional Markov models to estimate win probability throughout a curling match. In particular, the authors proposed both a homogeneous and heterogeneous Markov models for estimating the chance of winning in curling, with different independence assumptions on the relationship between performance and the current state of the game. In esports, @Maymin2020 used logistic regression to build a well-calibrated in-game win probability model for each specific moment during a game of League of Legends. More recently, @Guan2022 developed an in-game win probability model for the National Rugby League using functional data analysis. In this approach, the rugby event data are treated as functional, and the win probability is expressed as a function of the match time.

@@

Another important idea in sports analytics is the estimation of team strength. A popular method for estimating team strength in sports is through pairwise evaluations. Perhaps the most famous probability model for predicting the outcome of a paired comparison is the Bradley-Terry model (BTM) [@Bradley1952]. For a pair of players (or teams) $i$ and $j$, let $\Pi_{ij}$ denote the probability that $i$ is preferred to $j$. Then the BTM is a logistic regression model with parameters $\beta_i, \beta_j$ such that

@@

Here, $\exp{\beta_i}$ is often viewed as a representation of player $i$'s ability.

The BTM can be implemented in `R` [@Rcitation2022] via the \pkg{BradleyTerry2} package [@R-BradleyTerry2]. For example, we consider the data given in @agresti2018introduction, page 247 on tennis results from 2014 to 2018 for five men's professional players Djokovic, Federer, Murray, Nadal, and Wawrinka. We fitted a BTM to estimate the win probability for each pair of players and obtain a ranking for this group of five.

```{r}
tennis <- dplyr::tibble(
  Djokovic = c(NA, 9, 14, 9, 4),
  Federer = c(6, NA, 5, 5, 7),
  Murray = c(3, 0, NA, 2, 2),
  Nadal = c(2, 1, 4, NA, 4),
  Wawrinka = c(3, 2, 2, 3, NA)
) |>
  t() |>
  BradleyTerry2::countsToBinomial()

tennis_bt <- BradleyTerry2::BTm(cbind(win1, win2), player1, player2, data = tennis)

tennis_bt |>
  update(refcat = "Wawrinka") |> 
  BradleyTerry2::BTabilities() |> 
  round(3) |> 
  knitr::kable(caption = "Estimated abilities for each player, relative to Wawrinka, obtained from the fitted BTM. \\label{tab:tennisbtm}")
```

Table \ref{tab:tennisbtm} shows the estimated coefficients of the fitted BTM. According to the abilities, the players are ranked as follows: 1) Djokovic, 2) Federer, 3) Wawrinka, 4) Murray, 5) Nadal. To obtain win probability, as an illustration, for the Federer-Nadal matchup, an estimate for the probability of a Federer victory is
$$
\hat\Pi_{24} = \frac{exp(\hat\beta_2 - \hat\beta_4)}{1 + exp(\hat\beta_2 - \hat\beta_4)} = \frac{exp(1.136 + 0.062)}{1 + exp(1.136 + 0.062)} = 0.768
$$

```{r btm, eval=FALSE}
library(BradleyTerry2)
# show BTM for Pirates vs. Yankees in 1960
# or Red Sox vs. Yankees in 2003/2004
library(retrosheet)
games <- get_retrosheet("game", 2004)

@@

btm2 <- btm |>
  update(formula = ~ team + at_home, data = bt_games2)
btm2

btm |>
  broom::tidy() |>
  arrange(desc(estimate))
btm2 |>
  broom::tidy() |>
  arrange(desc(estimate))
```

Another widely known tool for measuring team strength is the Elo rating system [@Elo1978]. The Elo rating was originally developed for chess, but also has been used to estimate team strength in other sports. See @koning2017rating for more information on applications of the Elo rating in soccer.

@Glickman1998 proposed a Bayesian state-space model for paired comparisons for predicting NFL games, allowing team strength parameters to vary over time. In particular, this paper modeled point differential in the NFL by introducing week-to-week and season-to-season as the two primary sources variation in team strengths. See also @glickman2017estimating for more discussion on estimating team strengths in American football. More recently, @Lopez2018 extended @Glickman1998's state-space model to understand randomness in the four major American sports leagues. Betting moneylines were used in place of point differentials in order to estimate team strengths, and this framework also accounts for home advantage.
